---
title: "Getting Started with MAIVE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with MAIVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

MAIVE (Meta-Analysis Instrumental Variable Estimator) addresses a fundamental problem in meta-analysis of observational research: **spurious precision**. Traditional meta-analysis assigns more weight to studies with lower standard errors, assuming higher precision. However, in observational research, precision must be estimated and is vulnerable to manipulation through practices like p-hacking to achieve statistical significance.

For a guided, interactive workflow, visit <https://www.easymeta.org>.

This manipulation can invalidate:

- Inverse-variance weighting schemes
- Bias-correction methods like funnel plots
- Traditional publication bias corrections

MAIVE introduces an **instrumental variable approach** to limit bias caused by spurious precision in meta-analysis.

## The Problem: Spurious Precision

In observational research, researchers can inadvertently or deliberately manipulate their analyses to achieve statistically significant results. This includes:

- Selective reporting of specifications
- Outcome switching
- Sample trimming
- Selective controls inclusion

These practices create **spuriously precise estimates** that appear more reliable than they actually are. Traditional meta-analysis methods that weight by inverse variance will overweight these manipulated studies, leading to biased conclusions.

## The MAIVE Solution

MAIVE uses **instrumental variables** to correct for spurious precision:

1. **First-stage regression**: Instruments the potentially manipulated standard errors using sample sizes (which researchers cannot easily manipulate)
2. **Second-stage regression**: Uses the instrumented standard errors in meta-regression models

This approach provides:

- Robust meta-estimates that account for spurious precision
- Hausman-type tests comparing IV and OLS estimates
- Anderson-Rubin confidence intervals for weak instruments
- Publication bias tests based on instrumented standard errors

## Installation

```{r, eval = FALSE}
# Install from CRAN (once published)
install.packages("MAIVE")

# Or install development version from GitHub
install.packages("devtools")
devtools::install_github("PetrCala/MAIVE")
```

```{r setup}
library(MAIVE)
```

## Data Structure

The `maive()` function accepts either the default column names or custom mappings:

| Column              | Default name | Description                           |
|---------------------|--------------|---------------------------------------|
| Estimate            | `bs`         | Primary estimates (effect sizes)      |
| Std. Error          | `sebs`       | Standard errors (must be > 0)         |
| Sample size         | `Ns`         | Sample sizes (must be > 0)            |
| Study ID (optional) | `study_id`   | Clustering/fixed effects identifier   |

**Custom column names:** You can map your own column names using:

- `estimate` for the estimate column
- `se` for the standard error column
- `n` for the sample size column
- `study_id` for study identifiers (optional)

Example:

```{r custom-mapping}
custom_dat <- data.frame(
  my_est = c(0.50, 0.60, 0.40, 0.55, 0.45, 0.52, 0.48, 0.58, 0.42, 0.50, 0.47, 0.56),
  my_se = c(0.20, 0.18, 0.25, 0.22, 0.24, 0.19, 0.23, 0.17, 0.26, 0.21, 0.24, 0.18),
  my_n = c(80, 120, 95, 110, 90, 130, 100, 140, 85, 105, 88, 125),
  my_study = c("A", "A", "B", "B", "C", "C", "D", "D", "A", "B", "C", "D")
)

result <- maive(
  dat = custom_dat,
  estimate = "my_est",
  se = "my_se",
  n = "my_n",
  study_id = "my_study",
  method = 3,
  weight = 0,
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)

result$beta
```

Validation rules still apply: required columns must be numeric, non-missing, finite, and there must be at least 4 observations after removing completely empty rows. With study fixed effects (`studylevel` 1 or 3), there must be at least as many rows as unique studies plus three (the fixture above has 12 rows for 4 studies).

**Study identifier fallback:** when `study_id` is not supplied, a column named `study_id` is used wherever it sits. If there is no such column and the data frame has four or more columns, the fourth column is used as the study identifier and a warning names it. A column already mapped as `estimate`, `se`, or `n` is never picked: when column four is one of them, the single remaining column is used, and when several remain an error asks for `study_id` (at `studylevel = 0` the fallback is skipped, since no identifier is needed). If column four is a moderator or a year rather than a study identifier, drop it or name the correct column explicitly, because the identifier drives the study dummies and clustering at every `studylevel` other than 0. When `studylevel` is not 0 and there is no identifier at all, a warning says the study-level options have no effect.

**Coming from metafor:** if your effect sizes are already in a `metafor::escalc()` data frame or an `rma()` fit, `maive_from_metafor()` builds this data frame for you. It uses `sqrt(vi)` as the standard error and takes the sample sizes from the object (`ni`, the `ni` attribute that `escalc()` records, or `n1i + n2i`), never inferring them from the variance. The `ni` column and attribute are only used when they agree with `n1i + n2i` wherever both are present, because the attribute does not follow a reorder such as `dplyr::arrange()`. Rows of an `rma.uni` fit pass through the fit's `subset` and missing-value masks, so a `study_id` vector for the original data stays aligned. `metafor::trimfill()` fits are refused: their imputed studies would enter MAIVE as observed estimates, so pass the original `rma()` fit.

```{r metafor-bridge, eval = requireNamespace("metafor", quietly = TRUE)}
smd <- metafor::escalc(
  measure = "SMD",
  m1i = c(2.1, 2.5, 1.9, 2.4, 2.2, 2.6, 2.3, 2.0),
  sd1i = c(1.0, 1.2, 0.9, 1.1, 1.0, 1.3, 1.1, 1.0),
  n1i = c(40, 50, 30, 45, 60, 35, 55, 48),
  m2i = c(1.8, 2.0, 1.7, 2.1, 1.9, 2.2, 2.0, 1.8),
  sd2i = c(1.0, 1.1, 1.0, 1.0, 1.1, 1.2, 1.0, 1.1),
  n2i = c(42, 55, 33, 48, 58, 37, 52, 50)
)

maive_from_metafor(smd)
```

## Basic Usage

Let's create a simple example dataset:

```{r example-data}
# Simulated meta-analysis data: 50 estimates from 10 studies, with reported
# precision driven by sample size (as the first stage assumes)
set.seed(123)
n_estimates <- 50
Ns <- sample(100:1000, n_estimates, replace = TRUE)

data <- data.frame(
  bs = rnorm(n_estimates, mean = 0.3, sd = 0.2),
  sebs = 2 / sqrt(Ns) * runif(n_estimates, min = 0.8, max = 1.2),
  Ns = Ns,
  study_id = rep(1:10, each = 5)
)

head(data)
```

### Default MAIVE Estimation

The default MAIVE estimator uses PET-PEESE with instrumented standard errors, no weights, cluster-robust standard errors, and wild bootstrap:

```{r default-maive}
# Run MAIVE with defaults
result <- maive(
  dat = data,
  method = 3,      # PET-PEESE (default)
  weight = 0,      # No weights (default)
  instrument = 1,  # Instrument SEs (default)
  studylevel = 2,  # Cluster-robust (default)
  SE = 3,          # Wild bootstrap (default)
  AR = 1           # Anderson-Rubin CI (default)
)

# View key results
cat("MAIVE Estimate:", round(result$beta, 3), "\n")
cat("MAIVE SE:", round(result$SE, 3), "\n")
cat("Standard Estimate:", round(result$beta_standard, 3), "\n")
cat("Hausman Test:", round(result$Hausman, 3), "\n")
cat("First-stage F-test:", round(result$`F-test`, 3), "\n")
```

### Understanding the Output

The `maive()` function returns a list with the following key elements (see `?maive` for the full list):

- `beta`: MAIVE point estimate (corrected for spurious precision)
- `SE`: MAIVE standard error
- `beta_standard`, `SE_standard`: Conventional (non-IV) estimate and standard error for comparison
- `Hausman`: Hausman-type test comparing IV vs OLS estimates (high value suggests spurious precision)
- `F-test`: First-stage F-test of instrument strength
- `AR_CI`: Anderson-Rubin confidence interval (robust to weak instruments)
- `pub bias p-value`: p-value for the publication bias test based on the instrumented FAT
- `SE_instrumented`: Vector of instrumented standard errors, one per input row (`NA` for excluded estimates)
- `n_excluded`, `excluded_rows`: Number and positions of estimates excluded because the levels first stage fitted a non-positive variance for them (0 and `integer(0)` with the log first stage)
- `petpeese_selected`: Which of PET and PEESE was selected when `method = 3`
- `ek_structure`: Structure of the EK fit when `method = 4` ("kink", "linear", or "intercept")

## Method Options

MAIVE supports multiple meta-regression methods:

### 1. FAT-PET (Precision-Effect Test)

```{r pet, eval = FALSE}
result_pet <- maive(
  dat = data,
  method = 1,  # FAT-PET
  weight = 0,
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)

cat("PET Estimate:", round(result_pet$beta, 3), "\n")
```

### 2. PEESE (Precision-Effect Estimate with Standard Error)

```{r peese, eval = FALSE}
result_peese <- maive(
  dat = data,
  method = 2,  # PEESE
  weight = 0,
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)

cat("PEESE Estimate:", round(result_peese$beta, 3), "\n")
```

### 3. PET-PEESE (Conditional Method)

PET-PEESE uses PET if the PET estimate is not significantly different from zero, otherwise uses PEESE:

```{r petpeese, eval = FALSE}
result_petpeese <- maive(
  dat = data,
  method = 3,  # PET-PEESE (default)
  weight = 0,
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)

cat("PET-PEESE Estimate:", round(result_petpeese$beta, 3), "\n")
```

### 4. Endogenous Kink (EK)

```{r ek, eval = FALSE}
result_ek <- maive(
  dat = data,
  method = 4,  # EK
  weight = 0,
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 0  # AR not available for EK
)

cat("EK Estimate:", round(result_ek$beta, 3), "\n")
```

## Weighting Schemes

### No Weights (Default)

Unweighted regression, recommended when spurious precision is a concern:

```{r no-weights, eval = FALSE}
result_noweight <- maive(
  dat = data,
  method = 3,
  weight = 0,  # No weights
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)
```

### Inverse-Variance Weights

Traditional meta-analysis weighting:

```{r iv-weights, eval = FALSE}
result_ivweight <- maive(
  dat = data,
  method = 3,
  weight = 1,  # Inverse-variance weights
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 0  # AR not available with weights
)
```

### MAIVE-Adjusted Weights

Uses instrumented standard errors for weighting:

```{r maive-weights, eval = FALSE}
result_maiveweight <- maive(
  dat = data,
  method = 3,
  weight = 2,  # MAIVE-adjusted weights
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)
```

## Study-Level Correlation

Control for study-level correlation when you have multiple estimates per study:

```{r studylevel, eval = FALSE}
# No study-level adjustment
result_none <- maive(data, method = 3, weight = 0, instrument = 1, 
                     studylevel = 0, SE = 0, AR = 1)

# Study fixed effects (demeaned)
result_fe <- maive(data, method = 3, weight = 0, instrument = 1,
                   studylevel = 1, SE = 1, AR = 0)  # AR not available with FE

# Cluster-robust standard errors
result_cluster <- maive(data, method = 3, weight = 0, instrument = 1,
                        studylevel = 2, SE = 3, AR = 1)

# Both fixed effects and clustering
result_both <- maive(data, method = 3, weight = 0, instrument = 1,
                     studylevel = 3, SE = 3, AR = 0)
```

## Standard Error Options

```{r se-options, eval = FALSE}
# CR0 (Huber-White)
result_cr0 <- maive(data, method = 3, weight = 0, instrument = 1,
                    studylevel = 2, SE = 0, AR = 1)

# CR1 (Standard empirical correction)
result_cr1 <- maive(data, method = 3, weight = 0, instrument = 1,
                    studylevel = 2, SE = 1, AR = 1)

# CR2 (Bias-reduced estimator)
result_cr2 <- maive(data, method = 3, weight = 0, instrument = 1,
                    studylevel = 2, SE = 2, AR = 1)

# Wild bootstrap (recommended, default)
result_boot <- maive(data, method = 3, weight = 0, instrument = 1,
                     studylevel = 2, SE = 3, AR = 1)
```

## First-Stage Specification

MAIVE allows two functional forms for the first-stage regression. The log specification is the default; pass `first_stage = 0` to reproduce the levels specification from the published paper.

### Log (Default)

Log-linear regression of log(sebs²) on log(Ns) with smearing retransformation. The fitted variance is always positive, so no estimate can drop out of the second stage:

```{r first-stage-log, eval = FALSE}
result_log <- maive(data, method = 3, weight = 0, instrument = 1,
                    studylevel = 2, SE = 3, AR = 1, first_stage = 1)

cat("First-stage (log) F-test:", round(result_log$`F-test`, 3), "\n")
```

### Levels (Paper Specification)

Regresses variance (sebs²) on constant and 1/Ns. This is the specification in Irsova et al. (2025):

```{r first-stage-levels, eval = FALSE}
result_levels <- maive(data, method = 3, weight = 0, instrument = 1,
                       studylevel = 2, SE = 3, AR = 1, first_stage = 0)

cat("First-stage (levels) F-test:", round(result_levels$`F-test`, 3), "\n")
```

The levels fit is an ordinary least squares regression, so an individual fitted variance can be negative even when the intercept is not. Such an estimate has no instrumented standard error and is excluded from everything built on it (weights, PET, PEESE, PET-PEESE, EK, the F-test, the Hausman comparison, the Anderson-Rubin intervals, the bootstrap), so every reported statistic uses the same rows. A warning reports the count, which is also returned as `n_excluded` with the positions in `excluded_rows`. The log specification above, the default, cannot fit a negative variance and never excludes an estimate.

## WAIVE: More Aggressive Correction

WAIVE (Weighted Adjusted Instrumental Variable Estimator) provides a more aggressive correction for p-hacking and spurious precision by combining variance instrumentation with exponential-decay downweighting of studies with spurious precision or extreme outliers:

For technical details and methodology, see the [WAIVE slides](https://meta-analysis.cz/waive_ottawa.pdf).

```{r waive, eval = FALSE}
result_waive <- waive(
  dat = data,
  method = 3,
  weight = 0,
  instrument = 1,
  studylevel = 2,
  SE = 3,
  AR = 1
)

cat("WAIVE Estimate:", round(result_waive$beta, 3), "\n")
cat("WAIVE SE:", round(result_waive$SE, 3), "\n")
```

WAIVE is particularly useful when:

- You need a more aggressive correction for p-hacking beyond standard MAIVE
- You suspect extreme outliers in your data
- Standard errors may be severely manipulated
- You want automatic downweighting of both spuriously precise estimates and outliers

## Interpretation Guidelines

### Hausman Test

The Hausman test compares the MAIVE (IV) estimate with the standard (OLS) estimate:

- **High value**: Suggests spurious precision is a problem; MAIVE estimate is preferred
- **Low value**: IV and OLS are similar; spurious precision may not be severe

### First-Stage F-test

Tests the strength of the instrument (inverse sample size):

- **F > 10**: Strong instrument
- **F < 10**: Weak instrument; use Anderson-Rubin CI

### Anderson-Rubin Confidence Interval

Provides inference robust to weak instruments. Always check this CI when F-test is low.

### Publication Bias p-value

Tests for publication bias using instrumented FAT:

- **Low p-value**: Evidence of publication bias
- **High p-value**: Little evidence of publication bias

## References

Irsova, Z., Bom, P.R.D., Havranek, T., & Rachinger, H. (2025). Spurious precision in meta-analysis of observational research. *Nature Communications*, 16, 8454. [https://doi.org/10.1038/s41467-025-63261-0](https://doi.org/10.1038/s41467-025-63261-0)

Keane, M., & Neal, T. (2023). Instrument strength in IV estimation and inference: A guide to theory and practice. *Journal of Econometrics*, 235(2), 1625-1653.

## See Also

- `?maive` for detailed parameter documentation
- `?waive` for the robust WAIVE estimator
- Project website: <https://meta-analysis.cz/maive/>
